Proofs of Linearizability Based on Forward Simulations

### 2017-01-24 draftCAV/app-absImplStack.tex
dealing with simulations
queues

We have observed that some implementations (like time-stamped stack) do not have fixed remove (pop) linearization points that will correspond to $lin(pop,e,k)$ where $e$ could be a data value or $\texttt{EMPTY}$. However, we observe that, these implementations contain some points in their pop methods that logically removes the element from the pool. We call them commit points. If a method of a library has a fixed linearization  point, it is also a commit point. In this sense, commit points are weaker versions of fixed linearization points. Fixed linearization point of a pop preserves the following properties:
\begin{itemize}
\item If a $ret$ comes before a linearization point in the concrete execution, this order is preserved in the linearization of this execution.
\item If a $lin$ pop comes before another $lin$ pop in the concrete execution, this order is preserved in the linearization of this execution.
\item If a $lin$ pop comes before an $inv$ in the concrete execution, this order is preserved in the linearization of this execution. 
\end{itemize} 
A commit point is weaker than a fixed linearization point in the sense that it does not need to satisfy the first and the second conditions.

To our intuition, if a pop (remove) method has multiple finite linearization points, commit point is the latest element in this set. We have never observed an implementation in which a pop is linearized after it logically removes the element from the data structure.

From now on, we will restrict ourselves to stacks, since our example implementation that we will show linearizability of is the time-stamped (TS) stack. However, the notions and the machines we will introduce can be extended to queues easily.

We fix $\mathcal{M} = \{push, pop\}$ and $\mathcal{D} = \mathbb{N} \cup \{\texttt{EMPTY}\}$. We extend the alphabet $A\Sigma$ for stacks with commit points as $ACS\Sigma = A\Sigma \cup \{com(pop,d,k)|d \in \mathcal{D}, k \in \mathcal{M}\}$. We define cs-refinement and cs-linearizability as we defined q-refinement, q-linearizability in the previous sections. We also change definitions of backward and forward simulation relations for stacks with commit points as we do in the previous sections by replacing linearization points with commit points in this extensions. Lemma 1 of the first section still holds with new simulation relation definitions and cs-linearizability and cs-linearizability implies the original linearizability definition.

Our road map for this section is as follows: We will first introduce an intermediate stack machine $L_I$ that will be deterministic with respect to the alphabet $ACS\Sigma$. We will show that $L_I$ is equivalent to the standard abstract stack $L_A$ defined in the previous section with respect to the language $A\Sigma$. We show this by first showing that $L_I$ is a refinement of $L_A$ wrt alphabet $A\Sigma$ by finding a backward simulation relation between them and then $L_A$ is a refinement of $L_I$ with respect to alphabet $A\Sigma$ by finding a forward simulation relation between them. Since $L_I$ is deterministic wrt $ACS\Sigma$, if we have an implementation $L_C$ that is a cs-refinement of $L_I$, we can find a forward simulation between them. As an example, we will pick $L_C$ as time-stamped stack and establish a forward simulation relation between it and our $L_I$ machine.

Let us continue with defining $L_I$ first:
\item A state of $L_I$ again consists of a partial strict order and a program counter: $Q_I \subseteq ND \times ED \times (\mathbb{N} \rightarrow Lbl_I)$ where $Lbl_I = \{N, A_0, A_1, R_0, R_1, R_2\}$ is the set of transition labels for the operations. Different from the fixed linearization point case, this time nodes are not triples but 5-tuples of the form $(k,d,st,mc,con)$. First three fields are the same as previous intermediate machines: $k \in \mathbb{N}$ is operation identifier of a push operation, $d \in \mathcal{D}$ is the data value of that push and $st \in \{\texttt{PENDING}, \texttt{CLOSED}\}$ is the current status of the push operation. The fourth field $mc \subset \mathbb{N}$ keeps the operation identifiers of pop operations such that this push was maximally closed when the pop began. A node $n$ is maximally closed in a state $s$ iff $n.st  = \texttt{CLOSED}$ and if there is an edge $n \rightarrow n' \in ed_s$, then $n'.st = \texttt{OPEN}$. The fifth field $con \subset \mathbb{N}$ is the set of operation identifiers of the pop methods that are concurrent with this push i.e. the either this node was open when the pop started or this node is created when the pop was pending.
\item The transition labels consist of invocation and return actions of both methods and commit action for only pop method. Hence $\Sigma_I = ACS\Sigma$. Number of parameters for all actions common in previous intermediate stack machine and this intermediate stack machine are the same. Commit actions contain the second data field. They are of the form: $com(pop,d,k)$ where $d \in \mathcal{D}$ and $k \in \mathbb{N}$. 
\item Initial state consists of an empty strict partial order and a function mapping ever operation to $N$: ${q_0}_I = (\emptyset, \emptyset, f_{{q_0}_I})$ where  $f_{{q_0}_I}(i) =N$ for all $i \in \mathbb{N}$.
\item We define $\delta_I$ less formally, by giving verbal explanations to the transitions, omitting the obvious updates on the $f$ part and not mentioning about the parts of the nodes that does not change:
\item $(q, inv(push,d,k),q') \in \delta_I$ iff a new node $n=(k,d,\texttt{PENDING}, \emptyset, con_n)$ is added to $nd_{q'}$, where $con_n = \{ i \in \mathbb{N}| f_q(i) = R_0\}$; $n' \rightarrow n$ will be added to $ed_{q'}$ if $n'$ is a closed node at the state $q$.
\item $(q, ret(push,k),q') \in \delta_I$ iff either there is a \texttt{PENDING} node $n$ in state $q$ and this node becomes \texttt{CLOSED} in state $q'$ or there is no node with identifier $k$ in $q$ and nothing else than $f$ field changes in $q'$.
\item $(q, inv(pop,k),q') \in \delta_I$ iff for every open node $n \in nd_q$, $k$ is added to $n.con$ in $q'$ and for every maximally closed node $m \in nd_q$, $k$ is added to $m.mc$ in state $q'$. 
\item $(q, com(pop,k,d), q') \in \delta_I$ iff there exists a node $n$ in state $q$ such that $n.d = d$ and either $k \in n.con$ or $k \in n.mc$, this node $n$ and all the nodes adjacent to it are removed in the state $q'$, $k$ is removed from all $con$ and $mc$ fields of all nodes in state $q'$ and for all other pop operations $k'$ in $n.mc$ or in $n.con$ and for all states $n' \in nd_q$ such that $n' \rightarrow n \in ed_q$ and for all $n'' \in nd_q$, $n' \rightarrow n'' \in ed_q$ implies $k' \in n''.con$ (implies $k \notin n''.mc$ but it is stronger than this condition: if there are three closed states $p,q,r$ s.t. $p \rightarrow q$, $q \rightarrow r$, $p \rightarrow r$, $k'$ is only in $r.mc$ and we delete $r$, former condition only allows $k' \in q.mc$ whereas the latter one allows $k' \in p.mc$ in addition) we have $k' \in n'.mc$ in the state $q'$. Note that we need to assume data independence to make this action deterministic.
\item $(q, ret(pop,k,d), q') \in \delta_I$ iff $q=q'$ ignoring the $f$ fields.
$L_A$ is the same machine defined in the previous section. The common alphabet between $L_A$ and $L_I$ is $A\Sigma$. We will show that they are equivalent in terms of this alphabet.

\begin{lem}
$L_I$ is a refinement of $L_A$.
\end{lem}
\begin{proof}
We will provide a backward simulation relation $bs$ between states of $L_I$ and states of $L_A$. Our relation $bs \subseteq Q_I \times Q_A$ relates state $q=(so_q, f_q)$ to $q' =(s_{q'}, f_{q'}$ iff (i) for all operation identifiers $k \in \mathbb{N}$, if $f_q(k) \in \{N,R_1,R_2\}$, then $f_q(k) = f_{q'}(k)$; if $f_q(k) = A_0$, then $f_{q'}(k) = A_0$ and the data value $d$ associated with this add operation is not inserted to $s_{q'}$ or $f_{q'}(k) = A_1$;  if $f_q(k) = A_1$, then $f_{q'}(k) = A_2$; if $f_q(k) = R_0$, then $f_{q'}(k)=R_0$ or $f_{q'}(k)=R_1$; (ii) Let us call a pop operation pending if $f_{q}(k) = R_0$ and $PP_q$ be set of pending pops. There there exists a function $g: PP_q \rightarrow ND_q \cup \{\texttt{NONE}\}$ such that $k \in g(k).mc$ or $k \in g(k).con$ for all pop operation identifiers $k$ such that $g(k) \neq \texttt{NONE}$ and $g$ is one-to-one if we neglect \texttt{NONE}; $s_{q'}$ is obtained by extending $so_q$ to a total order in which pending nodes may not take place and nodes (open or closed) $n$ such that there exists a pop operation with identifier $k$ so that $g(k)=n$ surely do not take place, (iii) $g(k) = \texttt{NONE}$ implies $f_{q'}(k) = R_0$ and $g(k) \neq \texttt{NONE}$ implies $f_{q'}(k) =R_1$, (iv) if $g_{q'}(k) =n$ and $n$ is a pending node, then $f_{q'}(k') = A_1$ where $k'$ is the operation identifier part of $n$, (v) if there is a pending node with identifier $k$ and it takes part in the linearization, then $f_{q'}(k) = A_1$.

@@

Now, we show that TS-Stack is linearizable by showing the concrete TS-Stack implementation $L_C$ is a cs-refinement of $L_I$. As $L_C$, we pick the simplest version that omits the \texttt{EMPTY} returns of the pop  methods, does not allow unlinking and elimination.

>>>>

We show that $AbsS$ and $AbsS_0$ refine each other. The standard reference implementation $AbsS_0$ is defined exactly as the one for queues, $AbsQ_0$, except that pop linearization points extract values from the beginning of the sequence stored in the state.

\begin{figure} [t]
{\scriptsize
  \centering
  \begin{mathpar}
    \inferrule[call-push]{
      k\not\in dom(cp^0) \\ 
      d\neq {\tt EMPTY}
    }{
      \sigma,in^0,rv^0,cp^0
      \xrightarrow{inv(push,d,k)}
      %O\cup\{k\},<\cup \{(k',k): \ell_2(k')={\tt COMP}\},\ell[k\mapsto (d,{\tt PEND})],rv,cp[k\mapsto 1]
      \sigma,in^0[k\mapsto d], rv^0,cp^0[k\mapsto A_1]
    }\hspace{5mm}
    \inferrule[lin-push]{
      cp^0(k)=A_1
    }{
      \sigma,in^0,rv^0,cp^0
      \xrightarrow{lin(push,d,k)}
      %O\cup\{k\},<\cup \{(k',k): \ell_2(k')={\tt COMP}\},\ell[k\mapsto (d,{\tt PEND})],rv^0,cp[k\mapsto 1]
      d\cdot\sigma,in^0,rv^0,cp^0[k\mapsto A]
    }\hspace{5mm}

        \inferrule[ret-push]{
      cp^0(k)=A
    }{
      \sigma,in^0,rv^0,cp^0
      \xrightarrow{ret(push,k)}
      %O\cup\{k\},<\cup \{(k',k): \ell_2(k')={\tt COMP}\},\ell[k\mapsto (d,{\tt PEND})],rv^0,cp[k\mapsto 1]
      \sigma,in^0,rv^0,cp^0[k\mapsto A_2]
    }\hspace{5mm}

    \inferrule[call-pop]{
      k\not\in dom(cp^0) \\ 
    }{
      \sigma,in^0,rv^0,cp^0
      \xrightarrow{inv(pop,k)}
      %O\cup\{k\},<\cup \{(k',k): \ell_2(k')={\tt COMP}\},\ell[k\mapsto (d,{\tt PEND})],rv^0,cp[k\mapsto 1]
      \sigma,in^0,rv^0,cp^0[k\mapsto R_1]
    }\hspace{5mm}
        \inferrule[lin-pop1]{
      cp^0(k)=R_1 \\
      \sigma = d\cdot \sigma' 
    }{
      \sigma,in^0,rv^0,cp^0
      \xrightarrow{lin(pop,d,k)}
      %O\cup\{k\},<\cup \{(k',k): \ell_2(k')={\tt COMP}\},\ell[k\mapsto (d,{\tt PEND})],rv^0,cp[k\mapsto 1]
      \sigma',in^0,rv^0[k\mapsto d],cp^0[k\mapsto R_2]
    }\hspace{5mm}

        \inferrule[lin-pop2]{
      cp^0(k)=R_1 \\
      \sigma = \epsilon
    }{
      \sigma,in^0,rv^0,cp^0
      \xrightarrow{lin(pop,{\tt EMPTY},k)}
      %O\cup\{k\},<\cup \{(k',k): \ell_2(k')={\tt COMP}\},\ell[k\mapsto (d,{\tt PEND})],rv^0,cp[k\mapsto 1]
      \sigma,in^0,rv^0[k\mapsto {\tt EMPTY}],cp^0[k\mapsto R_2]
    }\hspace{5mm}
    \inferrule[ret-pop]{
      cp^0(k)=R_2 \\
      rv^0(k) = d
    }{
      \sigma,in^0,rv^0,cp^0
      \xrightarrow{ret(pop,d,k)}
      %O\cup\{k\},<\cup \{(k',k): \ell_2(k')={\tt COMP}\},\ell[k\mapsto (d,{\tt PEND})],rv^0,cp[k\mapsto 1]
      \sigma,in^0,rv^0,cp^0[k\mapsto R_3]
    }\hspace{5mm}
          \end{mathpar}
  }
 \vspace{-4mm}
  \caption{The transition relation of $AbsS_0$. 
  %\textcolor{red}{Call-Enq must have $d!= \texttt{EMPTY}$ as a premise. Also lin deq returning empty must be changed as before.}
  }
  \label{fig:transitions:AbsS_0}
\vspace{-2mm}
\end{figure}

Thus, the states of $AbsS_0$ are tuples $\tup{\sigma,in^0,rv^0,cp^0}$ where $\sigma\in\<Vals>^*$ is a sequence of values, $in^0:\<Ops> ~> \<Vals>$ records the input value of a push, $rv^0:\<Ops> ~> \<Vals>$ records the return value of a pop fixed at its linearization point ($~>$ denotes a partial function), and $cp^0:\<Ops> ~> \{A_1,A,A_2,R_1,R_2,R_3\}$ records the control point of every push ($A_1, A,A_2$) or pop operation ($R_1,R_2,R_3$).
All the components are $\emptyset$ in the initial state, and the transition relation $->$ is defined in Fig.~\ref{fig:transitions:AbsS_0}. The alphabet of $AbsS$ contains call/return actions and push/pop linearization points.

To prove that $AbsS$ is a refinement of $AbsS_0$ we define a normal $C\cup R$-backward simulation (i.e, a backward simulation as in Definition~\ref{def:back_app}) from $AbsS$ to $AbsS_0$. The reverse is shown using a normal $C\cup R$-forward simulation (i.e, a forward simulation as in Definition~\ref{def:for_app}).

%We have observed that some implementations (like time-stamped stack) do not have fixed remove (pop) linearization points that will correspond to $lin(pop,e,k)$ where $e$ could be a data value or $\texttt{EMPTY}$. However, we observe that, these implementations contain some points in their pop methods that logically removes the element from the pool. We call them commit points. If a method of a library has a fixed linearization  point, it is also a commit point. In this sense, commit points are weaker versions of fixed linearization points. Fixed linearization point of a pop preserves the following properties:
%\begin{itemize}
%\item If a $ret$ comes before a linearization point in the concrete execution, this order is preserved in the linearization of this execution.
%\item If a $lin$ pop comes before another $lin$ pop in the concrete execution, this order is preserved in the linearization of this execution.
%\item If a $lin$ pop comes before an $inv$ in the concrete execution, this order is preserved in the linearization of this execution. 
%\end{itemize} 
%A commit point is weaker than a fixed linearization point in the sense that it does not need to satisfy the first and the second conditions. 
%
%To our intuition, if a pop (remove) method has multiple finite linearization points, commit point is the latest element in this set. We have never observed an implementation in which a pop is linearized after it logically removes the element from the data structure. 
%
%From now on, we will restrict ourselves to stacks, since our example implementation that we will show linearizability of is the time-stamped (TS) stack. However, the notions and the machines we will introduce can be extended to queues easily. 
%
%We fix $\mathcal{M} = \{push, pop\}$ and $\mathcal{D} = \mathbb{N} \cup \{\texttt{EMPTY}\}$. We extend the alphabet $A\Sigma$ for stacks with commit points as $ACS\Sigma = A\Sigma \cup \{com(pop,d,k)|d \in \mathcal{D}, k \in \mathcal{M}\}$. We define cs-refinement and cs-linearizability as we defined q-refinement, q-linearizability in the previous sections. We also change definitions of backward and forward simulation relations for stacks with commit points as we do in the previous sections by replacing linearization points with commit points in this extensions. Lemma 1 of the first section still holds with new simulation relation definitions and cs-linearizability and cs-linearizability implies the original linearizability definition. 
%
%Our road map for this section is as follows: We will first introduce an intermediate stack machine $L_I$ that will be deterministic with respect to the alphabet $ACS\Sigma$. We will show that $L_I$ is equivalent to the standard abstract stack $L_A$ defined in the previous section with respect to the language $A\Sigma$. We show this by first showing that $L_I$ is a refinement of $L_A$ wrt alphabet $A\Sigma$ by finding a backward simulation relation between them and then $L_A$ is a refinement of $L_I$ with respect to alphabet $A\Sigma$ by finding a forward simulation relation between them. Since $L_I$ is deterministic wrt $ACS\Sigma$, if we have an implementation $L_C$ that is a cs-refinement of $L_I$, we can find a forward simulation between them. As an example, we will pick $L_C$ as time-stamped stack and establish a forward simulation relation between it and our $L_I$ machine. 
%
%Let us continue with defining $L_I$ first:
%\begin{itemize}
%\item A state of $L_I$ again consists of a partial strict order and a program counter: $Q_I \subseteq ND \times ED \times (\mathbb{N} \rightarrow Lbl_I)$ where $Lbl_I = \{N, A_0, A_1, R_0, R_1, R_2\}$ is the set of transition labels for the operations. Different from the fixed linearization point case, this time nodes are not triples but 5-tuples of the form $(k,d,st,mc,con)$. First three fields are the same as previous intermediate machines: $k \in \mathbb{N}$ is operation identifier of a push operation, $d \in \mathcal{D}$ is the data value of that push and $st \in \{\texttt{PENDING}, \texttt{CLOSED}\}$ is the current status of the push operation. The fourth field $mc \subset \mathbb{N}$ keeps the operation identifiers of pop operations such that this push was maximally closed when the pop began. A node $n$ is maximally closed in a state $s$ iff $n.st  = \texttt{CLOSED}$ and if there is an edge $n \rightarrow n' \in ed_s$, then $n'.st = \texttt{OPEN}$. The fifth field $con \subset \mathbb{N}$ is the set of operation identifiers of the pop methods that are concurrent with this push i.e. the either this node was open when the pop started or this node is created when the pop was pending.
%\item The transition labels consist of invocation and return actions of both methods and commit action for only pop method. Hence $\Sigma_I = ACS\Sigma$. Number of parameters for all actions common in previous intermediate stack machine and this intermediate stack machine are the same. Commit actions contain the second data field. They are of the form: $com(pop,d,k)$ where $d \in \mathcal{D}$ and $k \in \mathbb{N}$. 
%\item Initial state consists of an empty strict partial order and a function mapping ever operation to $N$: ${q_0}_I = (\emptyset, \emptyset, f_{{q_0}_I})$ where  $f_{{q_0}_I}(i) =N$ for all $i \in \mathbb{N}$.
%\item We define $\delta_I$ less formally, by giving verbal explanations to the transitions, omitting the obvious updates on the $f$ part and not mentioning about the parts of the nodes that does not change:
%\begin{itemize}
%\item $(q, inv(push,d,k),q') \in \delta_I$ iff a new node $n=(k,d,\texttt{PENDING}, \emptyset, con_n)$ is added to $nd_{q'}$, where $con_n = \{ i \in \mathbb{N}| f_q(i) = R_0\}$; $n' \rightarrow n$ will be added to $ed_{q'}$ if $n'$ is a closed node at the state $q$.
%\item $(q, ret(push,k),q') \in \delta_I$ iff either there is a \texttt{PENDING} node $n$ in state $q$ and this node becomes \texttt{CLOSED} in state $q'$ or there is no node with identifier $k$ in $q$ and nothing else than $f$ field changes in $q'$.
%\item $(q, inv(pop,k),q') \in \delta_I$ iff for every open node $n \in nd_q$, $k$ is added to $n.con$ in $q'$ and for every maximally closed node $m \in nd_q$, $k$ is added to $m.mc$ in state $q'$. 
%\item $(q, com(pop,k,d), q') \in \delta_I$ iff there exists a node $n$ in state $q$ such that $n.d = d$ and either $k \in n.con$ or $k \in n.mc$, this node $n$ and all the nodes adjacent to it are removed in the state $q'$, $k$ is removed from all $con$ and $mc$ fields of all nodes in state $q'$ and for all other pop operations $k'$ in $n.mc$ or in $n.con$ and for all states $n' \in nd_q$ such that $n' \rightarrow n \in ed_q$ and for all $n'' \in nd_q$, $n' \rightarrow n'' \in ed_q$ implies $k' \in n''.con$ (implies $k \notin n''.mc$ but it is stronger than this condition: if there are three closed states $p,q,r$ s.t. $p \rightarrow q$, $q \rightarrow r$, $p \rightarrow r$, $k'$ is only in $r.mc$ and we delete $r$, former condition only allows $k' \in q.mc$ whereas the latter one allows $k' \in p.mc$ in addition) we have $k' \in n'.mc$ in the state $q'$. Note that we need to assume data independence to make this action deterministic.
%\item $(q, ret(pop,k,d), q') \in \delta_I$ iff $q=q'$ ignoring the $f$ fields.
%\end{itemize}
%\end{itemize}
%$L_A$ is the same machine defined in the previous section. The common alphabet between $L_A$ and $L_I$ is $A\Sigma$. We will show that they are equivalent in terms of this alphabet.

\begin{lemma} 
$AbsS$ is a refinement of $AbsS_0$.
\end{lemma}
\begin{proof}
We define a normal $C\cup R$-backward simulation $bs$ from $AbsS$ to $AbsS_0$ as follows. Given an $AbsS$ state $s=\tup{O,<,\ell,rv,cp}$ and an $AbsS_0$ state $t=\tup{\sigma,in^0,rv^0,cp^0}$ we have that $(s,t)\in bs$ iff the following hold:
\begin{itemize}
	\item if a pop has committed or respectively, it has returned in $s$, then it had been linearized or respectively, it has returned in $t$, i.e., for every $k$, if $cp(k)\in\{R_2,R_3\}$ then $cp^0(k)=cp(k)$,
	\item a push is completed in $s$ whenever the same is true in $t$, i.e., for every $k$, $cp(k)=A_2$ iff $cp^0(k)=A_2$,
	\item a push is pending in $s$ iff either it is a non-linearized pending push in $t$ or its linearization point has been executed, i.e., for every $k$, $cp(k)=A_1$ iff $cp^0(k)=A_1$ and $in^0(k)$ doesn't occur in $\sigma$, or $cp(k)=A$,
	\item if a pop didn't commit in $s$ then it is pending in $t$ and it may have been linearized, i.e., for every $k$, if $cp(k)=R_1$ then $cp^0(k)\in \{R_1,R_2\}$,
	\item let ${\tt UNCOM\_POP}=\{k: cp(k)=R_1\}$ be the set of pending pops in $s$ that didn't commit. Then, there exists a function

	\item the sequence $\sigma$ is a linearization of a partial order $(D,\prec)$ where $D$ contains values labeling elements of $O$ and all the values corresponding to completed enqueues, i.e., $\ell_1({\tt COMP}(O))\subseteq D\subseteq \ell_1(O)$ ordered according to the happens-before order between the enqueues that added them, i.e., $d_1\prec d_2$ if{f} there exists $k_1,k_2$ such that $\ell_1(k_1)=d_1$, $\ell_1(k_2)=d_2$, and $k_1 < k_2$.
	\item the return values fixed at dequeue linearization points are the same, i.e., for every $k$, $rv(k)=rv^0(k)$,
	\item every dequeue is at the same control point in both $s$ and $t$, i.e., for every $k$ and $i\in \{1,2,3\}$, $cp(k)=R_i$ iff $cp^0(k)=R_i$,
	\item every pending enqueue has the same input value in both $s$ and $t$, i.e., for every $k$, $\ell_1(k)=in^0(k)$,
	\item a pending enqueue from $O$ has been linearized whenever its value is contained in $\sigma$, i.e., for every $k$, $cp^0(k)=A$ iff $\ell_1(k)\in D$ and $\ell_2(k)={\tt PEND}$, 
	\item a pending enqueue from $O$ hasn't been linearized whenever its value is not in $\sigma$, i.e., for every $k$, $cp^0(k)=A_1$ iff $\ell_1(k)\not\in D$ and $\ell_2(k)={\tt PEND}$, 
	\item a pending enqueue which is not in $O$ has been linearized, i.e., for every $k$, $cp^0(k)=A$ iff $k\not\in O$ and $cp(k)=A_1$, 
	\item an enqueue is completed in $s$ whenever it is completed in $t$, i.e., for every $k$, $cp(k)=A_2$ iff $cp^0(k)=A_2$,
\end{itemize}

In the following, we show that indeed $bs$ is a normal $C\cup R\cup Lin(deq)$-backward simulation from $AbsQ$ to $AbsQ_0$.

We will provide a backward simulation relation $bs$ between states of $L_I$ and states of $L_A$. Our relation $bs \subseteq Q_I \times Q_A$ relates state $q=(so_q, f_q)$ to $q' =(s_{q'}, f_{q'}$ iff (i) for all operation identifiers $k \in \mathbb{N}$, if $f_q(k) \in \{N,R_1,R_2\}$, then $f_q(k) = f_{q'}(k)$; if $f_q(k) = A_0$, then $f_{q'}(k) = A_0$ and the data value $d$ associated with this add operation is not inserted to $s_{q'}$ or $f_{q'}(k) = A_1$;  if $f_q(k) = A_1$, then $f_{q'}(k) = A_2$; if $f_q(k) = R_0$, then $f_{q'}(k)=R_0$ or $f_{q'}(k)=R_1$; (ii) Let us call a pop operation pending if $f_{q}(k) = R_0$ and $PP_q$ be set of pending pops. There there exists a function $g: PP_q \rightarrow ND_q \cup \{\texttt{NONE}\}$ such that $k \in g(k).mc$ or $k \in g(k).con$ for all pop operation identifiers $k$ such that $g(k) \neq \texttt{NONE}$ and $g$ is one-to-one if we neglect \texttt{NONE}; $s_{q'}$ is obtained by extending $so_q$ to a total order in which pending nodes may not take place and nodes (open or closed) $n$ such that there exists a pop operation with identifier $k$ so that $g(k)=n$ surely do not take place, (iii) $g(k) = \texttt{NONE}$ implies $f_{q'}(k) = R_0$ and $g(k) \neq \texttt{NONE}$ implies $f_{q'}(k) =R_1$, (iv) if $g_{q'}(k) =n$ and $n$ is a pending node, then $f_{q'}(k') = A_1$ where $k'$ is the operation identifier part of $n$, (v) if there is a pending node with identifier $k$ and it takes part in the linearization, then $f_{q'}(k) = A_1$.

@@
